### 2020-10-08 serial.tex
squoze dates back into abstract

The reproduction number \RR and the growth rate $r$ are critical quantities for characterizing outbreak dynamics.
They are linked by generation intervals, the time between infection and onward transmission.
Because generation intervals are difficult to observe, epidemiologists often substitute serial intervals, the time between symptom onset in successive links in a transmission chain.
Recent studies suggest that such substitution can bias estimates of \RR based on $r$.
Here we explore how these intervals vary over the course of an epidemic, and the implications for \RR estimation.
\emph{Forward-looking} serial intervals, measuring time forward from symptom onset of a focal individual, correctly describe the renewal process of symptomatic cases and therefore reliably link \RR with $r$.
In contrast, \emph{backward-looking} intervals, which measure time backward, and \emph{intrinsic} intervals, which neglect population-level dynamics, lead to incorrect \RR estimates.
Forward-looking intervals are affected both by epidemic dynamics and by censoring, changing in complex ways over the course of an epidemic.
We derive a heuristic method for addressing biases that arise from neglecting changes in serial intervals. 
We apply the method to early serial-interval-based estimates of  \RR for the COVID-19 outbreak in China outside Hubei province,
showing that using improperly defined serial intervals in this context can bias estimates of initial \RR by up to a factor of 2.6.
This study demonstrates the importance of early contact-tracing efforts and provides a framework for reassessing generation intervals, serial intervals, and \RR\ estimates for COVID-19.

>>>>

The reproduction number \RR and the growth rate $r$ are critical epidemiological quantities.
They are linked by generation intervals, the time between infection and onward transmission.
Because generation intervals are difficult to observe, epidemiologists often substitute serial intervals, the time between symptom onset in successive links in a transmission chain.
Recent studies suggest that such substitution biases estimates of \RR based on $r$.
Here we explore how these intervals vary over the course of an epidemic, and the implications for \RR estimation.
\emph{Forward-looking} serial intervals, measuring time forward from symptom onset of a focal individual, correctly describe the renewal process of symptomatic cases and therefore reliably link \RR with $r$.
In contrast, \emph{backward-looking} intervals, which measure time backward, and \emph{intrinsic} intervals, which neglect population-level dynamics, give incorrect \RR estimates.
Forward-looking intervals are affected both by epidemic dynamics and by censoring, changing in complex ways over the course of an epidemic.
We present a heuristic method for addressing biases that arise from neglecting changes in serial intervals. 
We apply the method to early (21 January 21 -- 8 February 8 2020) serial-interval-based estimates of  \RR for the COVID-19 outbreak in China outside Hubei province; using improperly defined serial intervals in this context biases estimates of initial \RR by up to a factor of 2.6.
This study demonstrates the importance of early contact-tracing efforts and provides a framework for reassessing generation intervals, serial intervals, and \RR\ estimates for COVID-19.
